\documentclass[12pt, oneside]{article}

\usepackage[margin=1.2in]{geometry}
\geometry{letterpaper}
\usepackage{tipa,vowel,gb4e}
\let\ipa\textipa
\let\eachwordone=\it

\usepackage{multirow}

\title{Constituent Order of the Ishiculu language}
\author{Ming Zhang}
\date{October 13, 2117}

\begin{document}
\maketitle

Some glossing abbreviations used in this chapter: \textsc{int} = interior, \textsc{sur} = surface, \textsc{static}, \textsc{prep} = preposition, \textsc{post} = postposition.

Ishiculu is subject-prominent. The basic word order of Ishiculu is SVO. More accurately, its basic word order is subject-predicate, where the predicate can be a VP, an AdjP, or a PP.

Ishiculu uses both prepositions and postpositions. All prepositions are used to mark locative case, but overt postpositions are only used in marking lative and ablative.

\begin{exe}
\ex
\gll Ari pa kiw-ioani nta\textbeltl o kitabu. \\
Ari not \textsc{3sg}-like \textsc{clf}.7 book(7) \\
\trans `Ari doesn't like the book.'
\ex
\gll Tse-kiw-ioani-mbi nta\textbeltl o kitabu. \\
\textsc{1pl}-\textsc{3sg}-like-\textsc{pst} \textsc{clf}.7 book(7) \\
\trans `We used to like the book.'
\ex
\gll Ari fu-Philadelphia-tso. \\
Ari \textsc{prep.int}-Philadelphia-\textsc{post.abl} \\
\trans `Ari is from Philadelphia.'
\ex
\gll {Fu-Philadelphia-\o} tse-ke-\textbeltl a-ki Ari nta\textbeltl o kitabu. \\
\textsc{prep.int}-Philadelphia-\textsc{post.sta} \textsc{1pl}-have-\textsc{caus}-\textsc{3sg} Ari \textsc{clf}.7 book(7) \\
\trans `We will give Ari the book in Philadelphia.'
\end{exe}

A locative phrase can also occupy the subject position of verb and the subject will be moved to the object position.

\begin{exe}
\ex
\gll Ts\textramshorns-nji-tso ki-huma nta\textbeltl o ika. \\
\textsc{prep.sur}-table-\textsc{post.abl} \textsc{3sg}-jump \textsc{clf}.6 cat(6) \\
\trans `Off the table jumps a cat.'
\ex
\gll {Fu-heje-\o} ki-paka i ntoch\textramshorns. \\
\textsc{prep.int}-courtyard-\textsc{post.abl} \textsc{3sg}-park \textsc{clf}.8 car(8) \\
\trans `In the courtyard is parked a car.'
\end{exe}

To put the emphasis on the object, it is possible to have OSV word order, but there will need to be a classifier for the object remaining in the original position.

\begin{exe}
\ex
\gll Kitabu tse-ki-so-mbi nta\textbeltl o. \\
book(7) \textsc{1pl}-\textsc{3sg}-read-\textsc{pst} \textsc{clf}.7 \\
\trans `The book, we have read it.'
\ex
\gll {ntoch\textramshorns} ni-kiw-ioani i. \\
car(8) \textsc{1sg}-\textsc{3sg}-like \textsc{clf}.8 \\
\trans `The car, I like it.'
\end{exe}

\end{document}